---
title: "P8105 HW2"
author: "Ruoqi Wang"
date: "2026-10-06"
output: github_document
---

## Problem 1

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r, message=FALSE, warning=FALSE}
transit_df = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("", "NA")
  ) |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    starts_with("route"),
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```

The NYC Transit dataset contains information about subway station entrances and exits in New York City, including the subway line and station name, station latitude and longitude, routes served by each station, whether entrance is allowed, vending availability, entrance type, and ADA compliance.I used read_csv() to read the local CSV; janitor::clean_names() to make the variable names easier to work with; select() to keep only the variables requested by the assignment; starts_with("route") to keep all the route variables; mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE)) to change entry from a character variable to a logical variable. The resulting dataset contains `r nrow(transit_df)` rows and `r ncol(transit_df)` columns. The data are not tidy yet because the routes served by each station are spread across multiple columns rather than being stored in a single route variable.

## How many distinct stations are there?
```{r}
transit_df |>
  distinct(line, station_name) |>
  nrow()
```

## How many stations are ADA compliant?
```{r}
transit_df |>
  distinct(line, station_name, ada) |>
  filter(ada == TRUE) |>
  nrow()
```

## What proportion of station entrances/exits without vending allow entrance?
```{r}
transit_df |>
  filter(vending == "NO") |>
  summarize(
    proportion_entry = mean(entry)
  )
```

## How many distinct stations serve the A train?
```{r}
transit_long = 
  transit_df |>
  mutate(
    across(route1:route11, as.character)
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  ) |>
  filter(!is.na(route_name))
```

```{r}
transit_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```

## Of the stations that serve the A train, how many are ADA compliant?
```{r}
transit_long |>
  filter(route_name == "A") |>
  distinct(line, station_name, ada) |>
  filter(ada == TRUE) |>
  nrow()
```


## Problem 2

```{r}
library(tidyverse)
library(readxl)
```

```{r}
mr_trash = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )
```

```{r, warning=FALSE}
professor_trash = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M144"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Professor Trash Wheel"
  )
```

```{r}
gwynnda_trash = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L349"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    trash_wheel = "Gwynnda"
  )
```

```{r}
trash_wheel_df = 
  bind_rows(
    mr_trash,
    professor_trash,
    gwynnda_trash
  )
```

The combined Trash Wheel dataset contains `r nrow(trash_wheel_df)` observations and `r ncol(trash_wheel_df)` variables. Each observation represents a dumpster collection from one of the three Trash Wheels: Mr. Trash Wheel, Professor Trash Wheel, or Gwynnda. Key variables include the Trash Wheel name, dumpster number, month, year, weight of trash in tons, volume of trash, and counts of various types of trash such as plastic bottles and cigarette butts. 

## Total weight of trash collected by Professor Trash Wheel.
```{r}
trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  )
```

## Total number of cigarette butts collected by Gwynnda in June of 2022.
```{r}
trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE)
  )
```


## Problem 3
```{r}
baseline_df = 
  read_csv(
    "data/MCI_baseline.csv",
    skip = 1,  #the file has a descriptive first row before the actual variable names
    na = ".",  #the file uses . to represent missing MCI-onset ages
    show_col_types = FALSE
  ) |>
  janitor::clean_names() |>
  mutate(
    sex = recode(  #removes anyone whose MCI onset age was at or before their baseline age
      sex,
      `0` = "Female",
      `1` = "Male"
    ),
    apoe4 = recode(  ##changes from numeric codes to meaningful carrier-status categories
      apoe4,
      `0` = "Non-carrier",
      `1` = "Carrier"
    )
  ) |>
  filter(  #removes anyone whose MCI onset age was at or before their baseline age
    is.na(age_at_onset) | age_at_onset > current_age
  )
```

## How many participants were recruited, and of these how many develop MCI?
```{r}
baseline_df |>
  summarize(
    recruited = n(),
    developed_mci = sum(!is.na(age_at_onset))
  )
```

## What is the baseline age?
```{r}
baseline_df |>
  summarize(
    mean_baseline_age = mean(current_age, na.rm = TRUE)
  )
```

## What proportion of women in the study are APOE4 carriers?
```{r}
baseline_df |>
  filter(sex == "Female") |>
  summarize(
    proportion_apoe4 = mean(apoe4 == "Carrier")
  )
```
After cleaning, the baseline dataset contains `r nrow(baseline_df)` participants. Of these participants, `r sum(!is.na(baseline_df$age_at_onset))` developed MCI during follow-up. The average baseline age was `r round(mean(baseline_df$current_age, na.rm = TRUE), 2)` years. Among women, `r scales::percent(mean(baseline_df$apoe4[baseline_df$sex == "Female"] == "Carrier"), accuracy = 1)` were APOE4 carriers.

```{r}
amyloid_df = 
  read_csv(
    "data/mci_amyloid.csv",
    skip = 1,  #the file has a descriptive first row before the actual variable names
    na = ".",
    show_col_types = FALSE
  ) |>
  janitor::clean_names() |>
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    values_to = "amyloid_ratio"
  )  #each row represents one amyloid measurement for one participant at one time point
```
The tidy amyloid dataset contains `r nrow(amyloid_df)` observations and `r ncol(amyloid_df)` variables, including participant ID, measurement time, and amyloid ratio.

```{r}
baseline_only = 
  baseline_df |>
  anti_join(
    amyloid_df,
    by = c("id" = "study_id")  #the baseline dataset uses the ID variable
  )

amyloid_only = 
  amyloid_df |>
  anti_join(
    baseline_df,
    by = c("study_id" = "id")  #the amyloid dataset uses study_id
  ) |>
  distinct(study_id)
```
Comparison of participant IDs showed that `r nrow(baseline_only)` participants appeared only in the cleaned baseline dataset, while `r nrow(amyloid_only)` participants appeared only in the amyloid dataset.

```{r}
combined_df = 
  baseline_df |>
  inner_join(  #it retains only IDs that occur in both datasets
    amyloid_df,
    by = c("id" = "study_id")
  )
write_csv(
  combined_df,
  "data/mci_amyloid_clean.csv"
)
```
The resulting combined dataset contains `r n_distinct(combined_df$id)` unique participants, `r nrow(combined_df)` observations, and `r ncol(combined_df)` variables. Each participant contributes longitudinal amyloid measurements along with their baseline demographic and MCI information.